---
title: "Technical report - Random Forest modeling (Forest floor C stocks)"
subtitle: "Machine learning Methodology"
author: "Dra. Zaira Rosario Pérez-Vázquez"
date: "Diciembre 2023"
output:
  html_document:
    toc: true
    number_sections: false
    toc_depth: 6
    fig_caption: true
    fig_path: "docs/"
    theme: paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Introduction

This technical report outlines the Random Forest modeling workflow used to map carbon stocks in the organic soil layers (forest floor: L and FH horizons). The process is divided into six stages, each documented in separate scripts located within the `/scripts` folder:

a) Exploratory Data Analysis
b) Feature Selection Procedure
c) Random Forest Modeling
d) Model Validation
e) Spatial Prediction of Carbon Stocks
f) Ecological Interpretation and Hotspot Mapping

# 2. Load requiered R packages 
These packages are required for data manipulation, statistical analysis, and plotting throughout the modeling workflow.


```{r libraries}
library(tidyverse)
library(psych)
library(here)
library(DescTools)
library(ggpubr)
library(car)
library(Hmisc)
library(moments)
library(cowplot)
library(readr)
```

# 3. Exploratory analysis (Stage 1)

## 3.1. Load data

**Important:**

- The file `RF_data_Carbon_ForestFloor.csv` is an example with a structure similar to the real data.
- The original data is not shared publicly for confidentiality reasons.
- The code was designed to work fully with this simulated version.

<div style="background:#f8f9fa; padding:10px; border-left:4px solid #007bff;">
<b>**🔒 Confidentiality Notice:**</b> 
If you have questions or need access to the real data, you can contact the repository author:

**Zaira Rosario Pérez-Vázquez**  
zairpv@gmail.com
</div>


```{r load-data}
data_path <- here("data", "RF_data_Carbon_ForestFloor.csv")
datasetFF <- read_csv(data_path)
str(datasetFF)
```

## 3.2. Data Preparation

The categorical variables are converted to factors to enable data grouping by sampling year, organic layer (L and FH), and forest management conditions.

```{r prepare}
datasetFF <- datasetFF %>%
  mutate(
    YEAR = as.factor(YEAR),
    LAYER = as.factor(LAYER),
    CONDITION = as.factor(CONDITION),
    LAYER_NUM = as.factor(LAYER_NUM),
    CON_NUM = as.factor(CON_NUM)
  )
```

## 3.3. Summary Statistics

To explore the distribution of carbon stocks in the forest floor, we computed descriptive statistics using the `psych` package.

- `describe()` provides overall statistics for the `C_STOCKS` variable, including mean, standard deviation, minimum, maximum, skewness, and kurtosis.
- `describeBy()` is used to calculate the same summary statistics, but grouped by **sampling year** and **organic layer** (L and FH), which helps us assess variation across both temporal and vertical gradients.
An

```{r descriptives}
describe(datasetFF$C_STOCKS)
describeBy(datasetFF$C_STOCKS, list(datasetFF$YEAR, datasetFF$LAYER))
```

## 3.4. Data distribution (Histograms and Normality)

To assess the distribution of carbon stocks (`C_STOCKS`) across sampling years (2013, 2018, 2023) and organic layers (L and FH), we used histograms and normality diagnostics.

Histograms were generated using `ggplot2`, displaying the frequency distributions of `C_STOCKS` faceted by year and layer. This layout facilitates the identification of temporal and vertical patterns in carbon accumulation.

**L layer**
```{r prepare_L_data, message=FALSE, warning=FALSE}
# Filtrar datos por año y capa L
data2013_L <- subset(datasetFF, YEAR == "2013" & LAYER == "L")
data2018_L <- subset(datasetFF, YEAR == "2018" & LAYER == "L")
data2023_L <- subset(datasetFF, YEAR == "2023" & LAYER == "L")
```

**FH layer**
```{r prepare_FH_data, message=FALSE, warning=FALSE}
# Filtrar datos por año y capa FH
data2013_FH <- subset(datasetFF, YEAR == "2013" & LAYER == "FH")
data2018_FH <- subset(datasetFF, YEAR == "2018" & LAYER == "FH")
data2023_FH <- subset(datasetFF, YEAR == "2023" & LAYER == "FH")
```

To evaluate the normality of carbon stock distributions in the **L layer**, we applied a combined visual and statistical approach. For each year, we produced histograms overlaid with both the kernel density estimate (blue line) and a theoretical normal distribution (red line), calculated using the observed mean and standard deviation. A dashed orange line indicates the mean value of each distribution.

### 3.4.1. Histograms - L layer
```{r histograms_L_panel_full, fig.width=12, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE}
# Expression for x-axis label
Mgha_expression1 <- expression("Mg ha"^"-1")

# Plot for 2013 - L
HIST_CL13 <- ggplot(data2013_L, aes(x = C_STOCKS)) +
  geom_histogram(aes(y=..density..), alpha=0.3, size=0.4, color="darkgrey", binwidth=0.5, fill="lightgray") +
  stat_function(aes(color="theoretical normal distribution"),
                fun=dnorm, args=list(mean=mean(data2013_L$C_STOCKS), sd=sd(data2013_L$C_STOCKS)), size=0.5) +
  geom_density(aes(color="density"), size=0.5) +
  geom_vline(aes(xintercept=mean(C_STOCKS, na.rm=TRUE), color="mean"), linetype="dashed", size=0.8) +
  scale_color_manual(name="Data distribution\n               L layer:",
                     values=c("density"="blue", "theoretical normal distribution"="red", "mean"="orange"),
                     labels=c("Observed density", "Mean value", "Theoretical normal")) +
  labs(x=Mgha_expression1, y="Density", title="2013") +
  annotate("text", x=4.85, y=0.68, label="p = 0.069", family="A") +
  theme_bw() +
  theme(text=element_text(size=12, family="A"),
        legend.position="bottom",
        legend.title=element_text(hjust=0.5, size=11, face="bold"),
        legend.text=element_text(size=11),
        plot.title=element_text(size=14, face="bold", hjust=0.5),
        axis.text=element_text(size=12))

# Plot for 2018 - L
HIST_CL18 <- ggplot(data2018_L, aes(x = C_STOCKS)) +
  geom_histogram(aes(y=..density..), alpha=0.3, size=0.4, color="darkgrey", binwidth=0.5, fill="lightgray") +
  stat_function(aes(color="theoretical normal distribution"),
                fun=dnorm, args=list(mean=mean(data2018_L$C_STOCKS), sd=sd(data2018_L$C_STOCKS)), size=0.5) +
  geom_density(aes(color="density"), size=0.5) +
  geom_vline(aes(xintercept=mean(C_STOCKS, na.rm=TRUE), color="mean"), linetype="dashed", size=0.8) +
  scale_color_manual(name="Data distribution\n               L layer:",
                     values=c("density"="blue", "theoretical normal distribution"="red", "mean"="orange"),
                     labels=c("Observed density", "Mean value", "Theoretical normal")) +
  labs(x=Mgha_expression1, y="Density", title="2018") +
  annotate("text", x=4.85, y=0.68, label="p = 0.069", family="A") +
  theme_bw() +
  theme(text=element_text(size=12, family="A"),
        legend.position="bottom",
        legend.title=element_text(hjust=0.5, size=11, face="bold"),
        legend.text=element_text(size=11),
        plot.title=element_text(size=14, face="bold", hjust=0.5),
        axis.text=element_text(size=12))

# Plot for 2023 - L
HIST_CL23 <- ggplot(data2023_L, aes(x = C_STOCKS)) +
  geom_histogram(aes(y=..density..), alpha=0.3, size=0.4, color="darkgrey", binwidth=0.5, fill="lightgray") +
  stat_function(aes(color="theoretical normal distribution"),
                fun=dnorm, args=list(mean=mean(data2023_L$C_STOCKS), sd=sd(data2023_L$C_STOCKS)), size=0.5) +
  geom_density(aes(color="density"), size=0.5) +
  geom_vline(aes(xintercept=mean(C_STOCKS, na.rm=TRUE), color="mean"), linetype="dashed", size=0.8) +
  scale_color_manual(name="Data distribution\n               L layer:",
                     values=c("density"="blue", "theoretical normal distribution"="red", "mean"="orange"),
                     labels=c("Observed density", "Mean value", "Theoretical normal")) +
  labs(x=Mgha_expression1, y="Density", title="2023") +
  annotate("text", x=4.85, y=0.68, label="p = 0.069", family="A") +
  theme_bw() +
  theme(text=element_text(size=12, family="A"),
        legend.position="bottom",
        legend.title=element_text(hjust=0.5, size=11, face="bold"),
        legend.text=element_text(size=11),
        plot.title=element_text(size=14, face="bold", hjust=0.5),
        axis.text=element_text(size=12))

# Combine the plots into a single panel
Supp1 <- ggarrange(HIST_CL13, HIST_CL18, HIST_CL23,
                   ncol=3, nrow=1,
                   common.legend=TRUE, legend="bottom")

# Display the panel
Supp1
```


### 3.4.2. Histograms - FH layer
```{r histograms_FH_panel_full, fig.width=12, fig.height=4, echo=TRUE, message=FALSE, warning=FALSE}

# Plot for 2013 - FH
HIST_CFH13=ggplot(data2013_FH, aes(x = C_STOCKS)) +
    geom_histogram(aes(y=..density..),alpha=0.3, size = 0.4, color = "darkgrey",binwidth = 4, fill = "lightgray", color = "black") +
    geom_density(color = "blue", size = 0.5,position = "identity") +
    labs(x = Mgha_expression1, y = "Density", title = "a) C stocks")+
    annotate("text", x=23.1, y=0.09, label= "p = 0.394",family = "A") +
    geom_vline(aes(xintercept=mean(C_STOCKS, na.rm=T)),   # Ignore NA values for mean
               color="orange", linetype="dashed", size=0.8)+
    stat_function(fun = dnorm, args = list(mean = mean(data2013_FH$C_STOCKS), 
                                           sd = sd(data2013_FH$C_STOCKS)), color = "red",  size = 0.5)+
    theme_bw()+
    theme(text = element_text(size=11,family = "A"),
          plot.title = element_text(size = 14,face = "bold",hjust = 0.5),
          plot.subtitle = element_text(size = 11,face = "bold",hjust = 0),
          axis.text = element_text(size = 11))+
    labs(title = "2013")

# Plot for 2018 - FH
HIST_CFH18=ggplot(data2018_FH, aes(x = C_STOCKS)) +
    geom_histogram(aes(y=..density..),alpha=0.3, size = 0.4, color = "darkgrey",binwidth = 1.5, fill = "lightgray", color = "black") +
    geom_density(color = "blue", size = 0.5,position = "identity") +
    labs(x = Mgha_expression1, y = "Density", title = "a) C stocks")+
    annotate("text", x=8.6, y=0.22, label= "p = 0.825",family = "A") +
    geom_vline(aes(xintercept=mean(C_STOCKS, na.rm=T)),   # Ignore NA values for mean
               color="orange", linetype="dashed", size=0.8)+
    stat_function(fun = dnorm, args = list(mean = mean(data2018_FH$C_STOCKS), 
                                           sd = sd(data2018_FH$C_STOCKS)), color = "red",  size = 0.5)+
    theme_bw()+
    theme(text = element_text(size=11,family = "A"),
          plot.title = element_text(size = 14,face = "bold",hjust = 0.5),
          plot.subtitle = element_text(size = 11,face = "bold",hjust = 0),
          axis.text = element_text(size = 11))+
    labs(title = "2018")

# Plot for 2023 - FH
HIST_CFH23=ggplot(data2023_FH, aes(x = C_STOCKS)) +
    geom_histogram(aes(y=..density..),alpha=0.3, size = 0.4, color = "darkgrey",binwidth = 2.5, fill = "lightgray", color = "black") +
    geom_density(color = "blue", size = 0.5,position = "identity") +
    labs(x = Mgha_expression1, y = "Density", title = "a) C stocks")+
    annotate("text", x=13, y=0.18, label= "p = 0.788",family = "A") +
    geom_vline(aes(xintercept=mean(C_STOCKS, na.rm=T)),   # Ignore NA values for mean
               color="orange", linetype="dashed", size=0.8)+
    stat_function(fun = dnorm, args = list(mean = mean(data2023_FH$C_STOCKS), 
                                           sd = sd(data2023_FH$C_STOCKS)), color = "red",  size = 0.5)+
    theme_bw()+
    theme(text = element_text(size=11,family = "A"),
          plot.title = element_text(size = 14,face = "bold",hjust = 0.5),
          plot.subtitle = element_text(size = 11,face = "bold",hjust = 0),
          axis.text = element_text(size = 11))+
    labs(title = "2023")

# Combine the plots into a single panel
Supp2=ggarrange(HIST_CFH13,HIST_CFH18,HIST_CFH23,
                ncol = 3,nrow = 1,
                common.legend = TRUE,legend="bottom")

# Display the panel
Supp2
```

### 3.4.3. Normality tests

Additionally, the **Kolmogorov-Smirnov test** was used to formally assess whether the data deviated significantly from a normal distribution. These results were visualized separately per year and integrated into a single panel for clarity.

In all three cases, the test returned **p-values greater than 0.5**, indicating that the null hypothesis of normality **could not be rejected**. These results suggest that the carbon stock data for the L layer is reasonably consistent with a normal distribution across the evaluated years.

**L layer**
```{r ks_test_2013_L, echo=TRUE, message=FALSE, warning=FALSE}
ks.test(data2013_L$C_STOCKS, 
        pnorm, 
        mean(data2013_L$C_STOCKS), 
        sd(data2013_L$C_STOCKS))
ks.test(data2018_L$C_STOCKS,pnorm,mean(data2018_L$C_STOCKS),sd(data2018_L$C_STOCKS))
ks.test(data2023_L$C_STOCKS,pnorm,mean(data2023_L$C_STOCKS),sd(data2023_L$C_STOCKS))
```

**FH layer**
```{r ks_test_2013_FH, echo=TRUE, message=FALSE, warning=FALSE}
ks.test(data2013_FH$C_STOCKS, 
        pnorm, 
        mean(data2013_FH$C_STOCKS), 
        sd(data2013_FH$C_STOCKS))
ks.test(data2018_FH$C_STOCKS,pnorm,mean(data2018_FH$C_STOCKS),sd(data2018_FH$C_STOCKS))
ks.test(data2023_FH$C_STOCKS,pnorm,mean(data2023_FH$C_STOCKS),sd(data2023_FH$C_STOCKS))
```


## 3.5. Spearman Correlations

To explore monotonic relationships between carbon stocks and a set of biotic and abiotic variables, we computed **Spearman’s rank correlation coefficients** for the year 2013. This non-parametric method is suitable for assessing correlations without assuming normality or linearity in the data.

The analysis included spatial coordinates (UTMX, UTMY), stand structure variables (STAND_AGE, BASAL_AREA, DOM_HEIGHT), biodiversity metrics (SPECIES_RICHNESS, SHANNON_INDEX), canopy features (GAP_FRACTION, CANOPY_COVER), and topographic attributes (ELEVATION, SLOPE, ASPECT).

The output provides:

- **Spearman's ρ (rho)** values, indicating the strength and direction of associations (from –1 to +1).
- **p-values**, indicating whether correlations are statistically significant.

Values close to ±1 represent strong associations. Significant positive or negative correlations with carbon stocks (`C_STOCKS`) suggest variables that may influence or co-vary with organic carbon accumulation in the forest floor during that year.

**Example for 2013**
```{r correlations}
data2013 <- datasetFF %>% filter(YEAR == "2013")
corr_vars <- data2013 %>% select(C_STOCKS, UTMX, UTMY, STAND_AGE, BASAL_AREA, DOM_HEIGHT, SPECIES_RICHNESS, SHANNON_INDEX, GAP_FRACTION, CANOPY_COVER, ELEVATION, SLOPE, ASPECT)
rcorr(as.matrix(corr_vars), type = "spearman")
```

## 3.6. Variance Inflation Factor (VIF)

To assess multicollinearity among predictor variables, we computed the **Variance Inflation Factor (VIF)** using a multiple linear regression model for the year 2013 as example. The model included spatial, structural, biodiversity, canopy, and topographic predictors associated with carbon stocks (`C_STOCKS`).

VIF quantifies how much the variance of a regression coefficient is inflated due to linear correlations with other predictors. It is used to detect redundancy in explanatory variables.

- A **VIF > 5** suggests moderate multicollinearity.
- A **VIF > 10** is generally considered problematic and may require variable removal or transformation.

The results help identify potential multicollinearity issues that could affect the stability and interpretability of model coefficients in subsequent modeling steps.

**Example for 2013**
```{r vif}
modelC_2013VIF <- lm(C_STOCKS ~ UTMX + UTMY + STAND_AGE + BASAL_AREA + DOM_HEIGHT + SPECIES_RICHNESS + SHANNON_INDEX + GAP_FRACTION + ELEVATION + SLOPE + ASPECT, data = data2013)
vif(modelC_2013VIF)
```

## 3.7. Train/Test Split
Before fitting the Random Forest models, we split the dataset into **training** and **validation** subsets to evaluate model performance on unseen data. This step is critical to avoid overfitting and ensure the generalizability of the model.

We used a random sampling approach for the year 2013, allocating:

- **85% of the data for training**, used to fit the model.
- **15% for validation**, used to assess prediction accuracy.

Stratification by organic layer (`LAYER`) was applied to maintain the proportion of L and FH samples across both subsets. This helps preserve the distributional characteristics of each layer during model training and evaluation.

**Example for 2013**
```{r split}
#set.seed(123)
#train_idx <- sample(seq_len(nrow(data2013)), size = 0.85 * nrow(data2013))
#training_2013 <- data2013[train_idx, ]
#testing_2013 <- data2013[-train_idx, ]

#write_csv(training_2013, here("data", "training_2013_85.csv"))
#write_csv(testing_2013, here("data", "testing_2013_15.csv"))
```

## 3.8. Conclusion

This exploratory analysis identified key variables, distributional patterns, and correlation structures to inform future Random Forest modeling stages. All outputs are saved in the `/outputs` folder, and this report is rendered to `/docs` for easy viewing.
